---
title: "Class 07: Machine Learning 1"
author: "Suraj Sidhu (A18512793)"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction

## K-Clustering

To see how this works, let's first make up some data to cluster where we know what the answer should be. We can use the `rnorm()` function to help here:

```{r}
hist(rnorm(100, mean = 5))
```


```{r}
x <- c(rnorm(30, mean=-3), rnorm(30, mean=3))
y <- rev(x)
```

*cbind binds by columns*
```{r}
x <- cbind(x,y) 
plot(x)
```

The function for k-Clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x, centers = 2)
k
```

To get at the results of the returned list object we can use the dollar `$` syntax

> Q. How many points are in each cluster ?

```{r}
k$size
```


> Q. What `component` of your result object details
- cluster assignment/membership ?
- cluster center ?

```{r}
k$cluster
```

```{r}
k$centers
```

> Q. Make a clustering results figure of the data colored by cluster membership and show cluster centers ?

```{r}
plot(x, col=c("forestgreen", "red"))
```

```{r}
plot(x, col=k$cluster, pch=16)
points(k$centers, col="blue", pch=15, cex=2)
```

K-means clustering is very popular as it is very fast and relatively straight forward: it takes numeric data as input and returns the cluster membership vector etc.

The "issue" is we tell `kmeans ()` how many clusters we want!

> Q. Run kmeans again and cluster into 4 groups/clusters and plot the results like we did above.

```{r}
k4 <- kmeans(x, centers =4)
plot(x, col=k4$cluster)
points(k4$centers, pch=15)
```
Scree plot to pick k `centers` value
```{r}
k1 <- kmeans(x, centers=1)
k2 <- kmeans(x, centers=2)
k3 <- kmeans(x, centers=3)
k4 <- kmeans(x, centers=4)
k5 <- kmeans(x, centers=5)
```

```{r}
c(k2$tot.withinss,
k3$tot.withinss,
k4$tot.withinss,
k5$tot.withinss)

plot(x, typ="b")
```

```{r}
n <- NULL
for(i in 1:5) {
 n <- c(n, kmeans(x, centers=i)$tot.withinss)
}
plot(n, typ="b")
```
## Hierarchical Clustering

The main "base" R function for Hierarchical Clustering is called `hclust()`. Here we can input our data we need to first calculate a distance matrix (e.g `dist()`) for our data and use this as a input to `hclust()`

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

There is a plot method for hclust results lets try it.

```{r}
plot(hc)
abline(h=8, col="red")
cutree(hc, h=8)
```

To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a given height or at a height that yields a given "k" groups

```{r}
cutree(hc, h=8)
```

```{r}
grps <- cutree(hc, k=2)
```

> Q. Plot the data with our hclust result coloring 

```{r}
plot(x, col=grps)
```


# Principle Component Analysis (PCA)

## PCA of UK food data

Import food data from a online CSV file:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```

```{r}
x <- read.csv(url, row.names=1)
x
```

Some base figures

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

```

There is one plot that can be useful for a small dataset:

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

> Main-Point: It can be difficult to spot major trends and patterns even in relatively small multi-variable datasets (here we only have 17 dimensions, typically we have thousands)


## PCA TO THE RESCUE

The main function in "base" R for PCA is called `prcomp()`
I will take the transpose of our data so the "foods" are in the columns:

```{r}
PCA <- prcomp(t(x))
summary(PCA)
```

```{r}
PCA$x
cols <- c("orange", "darkblue", "forestgreen", "red")
plot(PCA$x[,1], PCA$x[,2], col=cols, pch=16)
```

```{r}
library(ggplot2)

```

```{r}
ggplot(PCA$x) +
  aes(PC1, PC2) +
  geom_point(col=cols)
```

```{r}
ggplot(PCA$rotation) +
  aes(PC1, rownames(PCA$rotation)) +
  geom_col()
```




