Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
# Read the tab-delimited file
data <- read.table("rs8067378_ENSG00000172057.6.txt", header = TRUE)

# Check the first few rows
head(data)

```

```{r}
str(data)
summary(data)

```
```{r}
table(data$geno)
tapply(data$exp, data$geno, median)
```
```{r}
library(dplyr)

data %>%
  group_by(geno) %>%
  summarise(
    n = n(),
    median_expression = median(exp)
  )
```
Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
boxplot(exp ~ geno, data = data,
        main = "ORMDL3 Expression by rs8067378 Genotype",
        xlab = "Genotype",
        ylab = "ORMDL3 Expression",
        col = c("lightblue", "lightgreen", "pink"))

```
The boxplot shows that median ORMDL3 expression varies among genotypes. The G/G genotype exhibits higher median expression compared to the A/A genotype, indicating that the rs8067378 SNP is associated with differences in ORMDL3 expression, with the G allele linked to increased expression.